\section{Conclusion}

As shown in~\cite{guarnieri2016strong}, current database access control mechanisms are not secure.
%
Therefore, databases need new, stronger access control mechanisms to prevent attacks threatening both database integrity and confidentiality. %is here and database vendors need to address this fact in the future.
%
In this thesis, we address two main issues that 
%
First, we refactor the code of a modern open-source DBMS, namely PostgreSQL, in such a way to ease the implementation of new access control mechanisms.
%
The current PostgreSQL's architecture is not adequate to support these new mechanisms, since access control is closely tied with the database internals.
%
We performed  major modifications to the PostgreSQL's code addressing access control.
%
\remark{A few more information here?}
%
These modifications pave the way for the implementation of new access control mechanisms by providing developers with a uniform interface and clear code segregation.


Second, we developed an algorithm for deciding whether queries are secure or not.
%
Our algorithm directly applies to the SQL language.
%
As shown also in~\cite{guarnieri2016strong}, this kind of algorithms is an important building block for novel access control mechanisms that aim at prevent the leakage of sensitive information.
%
Existing algorithms, such as those proposed by the SQL standard and implemented in PostgreSQL, are overly restrictive.
%
Using them as building blocks for novel mechanisms would result in almost unusable systems.
%
Our solution, in contrast, exploits advanced query rewriting techniques to determine whether queries are secure or not. 
%
As a result, it is more permissive than existing solutions.

\remark{Conclusion should also discuss future work/possible extension/ lesson learned. How do you think other people will use what you've done.}


\remark{Most of this discussion should probably go in the related work section.}
The algorithms devised in this thesis are constructed under the \emph{Non-truman model}.
%
It specifies an access control mechanism as a component which decides if a query is optimized or not, and therefore returns $\bot$ or $\top$.
%
The current implementations of these mechanisms either interrupt the execution of the query and notify the user if a query was not authorized or do not interrupt the execution of the query if the query was authorized.
%
The Non-truman model is the most used model in database access control mechanisms.
%
The \emph{Truman model} however uses access control mechanisms with another approach, namely the access control internally executes the query, and only returns the tuples, which are readable for the user according to the access control policy.
%
The user issuing a query will not notice if the query he issued is not authorized for him.
%
Our algorithm is extensible to, and even would be better suited for the Truman model.
%
We under-approximate the tuples resulting from a query during the algorithm that returns the access control decision, and we further know that the under-approximation is secure with respect to the access control policy.
%
Therefore it is secure to return the under-approximation to the user.
%
Our implementation contains a subset of the introduced inference rules for containment, implementing all of them would result in a better under-approximation.

\remark{on what do we want to focus further here?}